Statistical Inference for inter-arrival times of extreme events in bursty time series

### 2020-07-27 article_springer/article_springer.Rmd
Write new section "Mittag-Leffler Stability Plots"

# Continuous Time Random Exceedances (CTRE)

@@

# Model Choice and inference for the Mittag-Leffler distribution{#sec:ML}

The classical POT approach is based on the fact that exceedances above a high threshold are asymptotically GPD distributed, hence a GPD distribution is fitted to the exceedances above a high enough threshold. As this approach is by now well-established in extreme value theory, we do not discuss it any further. Instead, this article focuses on our new contribution: the modelling of inter-exceedance times. For literature pointing to POT see e.g. @smith1984threshold; @leadbetter1991basis; @davison1990models; @beirlantBook Ch.5.3; @embrechts2013modelling Ch. 6.5.; @ColesBook Ch.4.

Due to the Theorem in the last Section we know that for high thresholds the times between exceedances are asymptotically Mittag-Leffler distributed in case of heavy-tailed waiting times between the observations. In this article we propose to model inter-exceedance times in a similar fashion as magnitudes in the classical POT approach, more precisely to fit a Mittag-Leffler distribution to the inter-exceedance times via stability plots. Hence, before we discuss inference for the inter-exceedance times in the next section, we first discuss inference for Mittag-Leffler distributions.

>>>>

# Continuous Time Random Exceedances (CTRE) {#sec:CTRE}

@@

# Mittag-Leffler Stability Plots

We assume a dataset of the form $\{(T_i, X_i), i \in I\}$,
where $T_i$ are timestamps and $X_i$ are magnitudes.
In this article, we focus on modelling the distribution of threshold
*exceedance times* rather than the *exceedances* themselves. For the latter, 
see the POT approach [@smith1984threshold;
@leadbetter1991basis; @davison1990models; @beirlantBook Ch.5.3;
@embrechts2013modelling Ch. 6.5.; @ColesBook Ch.4.].
With the previous section setting the stage, we assume that only the large
magnitudes (aka "shocks") follow a MRP.  The smaller magnitudes are possibly
nuisance data that may be irrelevant with respect to the modelling of extremes
and their waiting times. 
Accordingly, we assume a threshold $\ell$, and discard all
data $(T_i, X_i)$ where $X_i \le \ell$, yielding the thinned dataset
$\{(T_i, X_i), i \in I(\ell)\}$, where $I(\ell) \subseteq I$.

However, this raises the question of how high or low the threshold should be
chosen. Recall that with the POT method for the modelling of extremes, the
distribution of the thresholded exceedances converges to a GPD, and the
threshold must be chosen high enough so that the exceedances fit a GPD well. 
The threshold choice translates into a
bias-variance trade-off: low thresholds yield more data to fit a GPD (low 
variance), but the distribution of exceedances may deviate from a GPD 
distribution (high bias); and high thresholds present the opposite scenario.
It is understood, however, that there is a *range* of threshold values that
ought to yield similar GPD parameter estimates. In other words, plots of
threshold vs. parameter estimate ought to exhibit that parameter estimates
are *stable* with respect to the choice of threshold.

For inter-exceedance *times* (IETs), the same idea applies. 
By the Theorem, the inter-exceedance times converge to a Mittag-Leffler
distribution (MLD), and importantly, the tail parameter $\beta$ is 
*independent* of (or *stable* with respect to) the choice of threshold.
For high thresholds, the IET distribution may fit an MLD well, but few
IETs may be left for a fit (high variance). 
For low thresholds, their distribution may be not well represented
by an MLD; or, too many events may be interpreted as 
shocks, resulting in a biased estimate of the parameters of the limiting MLD.
For illustration, consider the case where measurements are taken daily:
Then at the lowest threshold, all IETs equal one day,
leaving no hint at a power-law distribution whatsoever. Raising the threshold
may or may not produce IETs that start exhibiting a power-law; and both cases
are represented by the MLD, with $\beta < 1$ or $\beta = 1$.

Suppose the "correct" threshold $\ell_0$ was known, in the sense that 
thresholding at $\ell_0$ would result precisely in an underlying MRP 
$\{(W_i, J_i): i \in I\}$ as defined in Section \ref{sec:CTRE}. 
In this case, we could apply any tail (and scale) estimator such as e.g.
Hill's estimator to the $W_i$ and thus infer the distribution of 
inter-exceedance times.

But guessing the correct threshold $\ell_0$ and proceeding with this
type of inference is less favourable to modelling IETs with the MLD, 
as we now explain.
In real-world datasets, a "correct" threshold may not exist: the data 
$\{(T_i, X_i): i \in I\}$ may be a mixture of an MRP with some other
time series $\{(T'_i, X'_i): i \in I'\}$, and the supports $\mathcal S'$
and $\mathcal S$ of the distributions
of $X'_i$ and the magnitudes $J_i$ (aka the MRP marks) may overlap.
Then thresholding at e.g. $\ell_0 = \sup \mathcal S'$ 
no longer results in a "clean" MRP: some $J_i$'s will be removed, meaning
that some $W_i$'s will be merged into a longer IET.
The more the threshold $\ell_0$ cuts into $\mathcal S$ from below, the more
the distributions of IETs will resemble a MLD, according to the Theorem.
As we do not know to what extent this effect occurs, 
and since the limit is an MLD, it is practicable to fit an MLD to the IETs to 
begin with.

# Fitting Mittag-Leffler Distributions {#sec:ML}
